Goal Recognition as Planning with Unknown Domain Models

### 2019-01-31 doc/ijcai-19-goalRecognition/goalRecognition.tex
first version of the ijcai19 paper on goal recognition with unknown domain models (no experiments)

\begin{abstract}
This paper shows how to relax a strong assumption of the {\em plan recognition as planning} approach that is {\em knowing the action model of the observed agents}. The paper introduces a novel formulation for classical planning in a setting where no action model is given (instead, only the state variables and the action headers are known) and it shows that this formulation neatly fits with the popular {\em plan recognition as planning} approach for {\em goal recognition}. The empirical evaluation evidences that this novel formulation allows to solve standard goal recognition benchmarks, still using an off-the-shelf classical planner, but without {\em a priori} knowing the action model of the observed agents.  
\end{abstract}

\section{Introduction}
\label{sec:introduction}
{\em Goal recognition} is a particular classification task in which each class represents a different goal and classification examples are observations of agents acting to achieve one of that goals. Despite there exists a wide range of different approaches for {\em goal recognition}, {\em plan recognition as planning}~\cite{ramirez2009plan,ramirez2012plan} is one of the most appealing and it is currently at the core of various activity recognition tasks such as, {\em goal recognition design}~\cite{KerenGK14}, {\em deceptive planning}~\cite{masters2017deceptive}, {\em planning for transparency}~\cite{macnally2018action} or {\em counter-planning}~\cite{PozancoEFB18}.

{\em Plan recognition as planning} leverages the action model of the observed agents and an off-the-shelf classical planner to compute the most likely goal of that agents. In this paper we show how to relax the assumption of {\em knowing the action model of the observed agents}, which frequently becomes a too strong assumption when applying {\em plan recognition as planning} at real-world problems.  In particular, the paper introduces a novel formulation for classical planning in a setting where no action model is given (instead, only the state variables and the action headers are known) and it shows that this formulation neatly fits with the successful {\em plan recognition as planning} approach. The empirical evaluation evidences that this novel formulation allows to solve standard goal recognition benchmarks, still using an off-the-shelf classical planner, but without {\em a priori} knowing the action model of the observed agents.

\section{Background}
\label{sec:background}
This section formalizes the {\em planning model} we follow, the kind of {\em observations} given as input to the {\em goal recognition} task, and the {\em plan recognition as planning} approach for {\em goal recognition}.

@@

\begin{itemize}
\item The {\bf observed actions} are consistent with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. Specifically, the number of observed actions, $l$, can range from $0$ (fully unobserved action sequence) to $|\pi|$ (fully observed action sequence).
\item The {\bf observed states} $\tup{s_0^o, s_1^o, \ldots, s_m^o}$ is a sequence of possibly {\em partially observable states}, except for the initial state $s_0^o$, which is fully observed. A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,P)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, range from 1 (the initial state, at least), to $|\pi|+1$, and each observed states comprises $[1,|F|]$ fluents.
\end{itemize}

@@

\subsection{Goal recognition with classical planning}
{\em Goal recognition} is a particular classification task in which each class represents a different goal $G\in G[\cdot]$ and there is a single classification example, $\mathcal{O}(\tau)$, that represents the observation of agents acting to achieve one of the input goals in $G\in G[\cdot]$. Following the {\em naive Bayes classifier}, the {\em solution} to the {\em goal recognition} task is the subset of goals in $G[\cdot]$ that maximizes this expression.
\begin{align}
argmax_{G\in G[\cdot]} P(\mathcal{O}|G) P(G).
\end{align}

The {\em plan recognition as planning} approach shows that the $P(\mathcal{O}|G)$ likelihood can be estimated leveraging the action model of the observed agents and an off-the-shelf classical planner~\cite{ramirez2012plan}. Given a {\em classical planning problem} $P=\tup{F,A,I,G[\cdot]}$ (where $G[\cdot]$ represents the set of possible goals) then $P(\mathcal{O}|G)$ is estimated computing, for each goal $G\in G[\cdot]$, the cost difference of the solution plans to these two different classical planning problems:
\begin{itemize}
\item $P^{\top}_G$, the classical planning problem built constraining $P=\tup{F,A,I,G}$ to achieve the particular goal $G\in G[\cdot]$ through a plan $\pi^\top$ that is {\em consistent} with the input observation $\mathcal{O}(\tau)$.
\item $P^{\bot}_G$, the classical planning problem that constrains the solutions of $P=\tup{F,A,I,G}$ to plans $\pi^\bot$ that achieve $G\in G[\cdot]$ but are {\em inconsistent} with $\mathcal{O}(\tau)$.
\end{itemize}

The higher the value of the $cost(\pi^\top)-cost(\pi^\bot)$ difference, the higher is the probability of the observed agents to aim goal $G\in G[\cdot]$. With this regard, {\em plan recognition as planning} uses the {\em sigmoid function} to map the previous cost difference into a likelihood:

@@

\section{Planning with unknown domain models}
\label{sec:planning}
This section introduces a novel formulation for classical planning in a setting where no action model is given. This setting of classical planning has already shown related to the learning of action models for planning~\cite{SternJ17}. In particular it can be seen an extreme scenario when the action model is learned from a single example that contains only two state observations: an initial state and the goals. A {\em classical planning with unknown domain models} is then a tuple $P=\tup{F,A[\cdot],I,G}$, where $A[\cdot]$ is a set of actions s.t., the semantics of each action $a\in A[\cdot]$ is unknown (i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined).

A solution to this task is a sequence of actions $\pi=\tup{a_1, \ldots, a_n}$ whose execution induces a trajectory $\tau(\pi,I)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and {\em there exists} at least one possible action model (e.g. one possible definition of the $\rho$ and $\theta$ functions within the given state variables) that satisfies $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$, for every {\small $1\leq i\leq n$}, and such that the reached final state met the goal conditions, $G \subseteq s_n$.

Next we show that the space of possible STRIPS action models can be encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this encoding to solve $P=\tup{F,A[\cdot],I,G}$ problems with an off-the-shelf classical planner as well as the properties of this approach.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, the actual space of possible \strips\ schemata is bounded by constraints of three kinds:
\begin{enumerate}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are also constraints of this kind~\cite{fox1998automatic}. 
\item {\em Observation constraints}. An observations $\mathcal{O}(\tau)$ depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

In this work we introduce a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of a \strips\ action schema $\xi$ using only fluents of two kinds {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$). This encoding exploits the syntactic constraints of \strips\, so it is more compact that the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} for learning classical planning action models. In more detail, if {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em negative effect} in $\xi$ while if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em positive effect} in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema with {\tt\small pre\_e\_stack} and {\tt\small eff\_e\_stack} fluents ($e\in{\mathcal I}_{\Psi,stack}$).

\subsection{A classical planning compilation for planning with unknown domain models}
Now we show how we adapt the classical planning compilation for learning \strips\ action models~\cite{aineto2018learning} for addressing the task of planning with unknown domain models with our propositional encoding of STRIPS action models.

Given a classical planning problem with unknown domain models $P=\tup{F,A[\cdot],I,G}$ we create a classical planning problem $P'=\tup{F',A',I,G}$ such that:
\begin{itemize}
\item $F'$ extends $F$ with the fluents for the propositional encoding of the corresponding space of STRIPS action models. This is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding action parameters $pars(\xi)$.

\item $A'$ replaces the actions in $A$ with two types of actions.
\begin{enumerate}
\item Actions for {\em inserting} a component (precondition, positive effect or negative effect) in $\xi$ following the syntactic constraints of \strips\ models. 
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in \Psi_{\xi}$ to the action model $\xi$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$, $eff_p$ exist in $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p}(\xi), \neg eff_{p}(\xi)\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p}(\xi)\}.
\end{align*}
\end{small}

@@

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p}(\xi)\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}

\item Actions for {\em applying} an action model $\xi$ built by the insert actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects to the $\Psi$ predicates and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). Since action headers are known, $pars(\xi)$ are bound to the objects in $\omega$ that appear in the same position.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\implies p(\omega)\}_{\forall p\in\Psi_\xi},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{\neg pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi}\}.
\end{align*}
\end{small}

\end{enumerate}
\end{itemize}

The intuition of the compilation is that the dynamics of the actions for {\em applying} an action model $\xi$ is determined by the values of the corresponding fluents $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ in the current state. For instance, executing {\tt{\small (apply\_stack blockB blockA)}} in a state $s$ implies activating the preconditions and effects of {\tt{\small (apply\_stack)}} according to the values of $\{pre\_e\_stack, eff\_e\_stack\}_{\forall e\in{\mathcal I}_{\Psi,stack}}$ fluents in $s$. This means that if the current state $s$ holds $\{{\tt\scriptsize (pre\_stack\_holding\_v1),(pre\_stack\_clear\_v2)}\} \subset s$, then it must be checked that positive literals {\tt{\small (holding blockB)}} and {\tt{\small (clear blockA)}} hold in $s$. Otherwise, a different set of precondition literals will be checked for the stack action. The same applies to the positive and negative effects. Executing {\tt{\small (apply\_stack blockB blockA)}}, will add the literals {\tt{\small (on blockB blockA),(clear blockB),(not(clear blockA)),(handempty)}} and {\tt{\small(not(clear blockB))}} to the successor state only if {\tt{\small stack}} has been correctly programmed by the {\em insert} actions.

\begin{figure}[hbt!]
	{\scriptsize\tt
		{\bf 00} : (insert\_pre\_stack\_holding\_v1) \\
		01 : (insert\_pre\_stack\_clear\_v2)\\
		{\bf 02} : (insert\_eff\_stack\_clear\_v1)\\
		03 : (insert\_eff\_stack\_clear\_v2)\\
		04 : (insert\_eff\_stack\_handempty)\\
		05 : (insert\_eff\_stack\_holding\_v1)\\
		06 : (insert\_eff\_stack\_on\_v1\_v2)\\
		{\bf 07} : (apply\_stack blockA blockB)\\
	}
	\caption{\small Plan computed when solving the classical planning problem output by our compilation for solving a classical planning with unknown domain models.}
	\label{fig:plan-lplan}
\end{figure}

Figure~\ref{fig:plan-lplan} shows a solution plan computed when solving a $P'=\tup{F',A',I,G}$ classical planning problem output by our compilation. In the initial state of that problem the robot is holding {\tt blockA} while {\tt blockB} is clear and the single problem goal is having {\tt blockA} on top of {\tt blockB}. The plan shows that the {\em insert} actions for the action model {\tt\small stack} (steps $00-01$ insert the preconditions of the {\tt\small stack} model, steps $02-06$ insert the action model effects), and step $07$ is the plan postfix that applies the programmed action model to achieve the goals $G$. Note that another valid solution could be computed for instance by inserting the same preconditions and effects into the {\em unstack} action model and then applying action {\tt\small (unstack blockA blockB)} instead of {\tt\small(stack blockA blockB)}.

\subsection{Compilation properties}
Now we present the properties of the compilation scheme.

@@

\begin{proof}[Proof]
\begin{small}
Once a given precondition or effect is inserted into an action model it can never be removed back and once an action model is applied it cannot be {\em reprogramed}. In the compiled problem the value of $F$, the fluents of the original problem, can exclusively be modified via $\mathsf{apply_{\xi,\omega}}$ actions.  The set of goals $G$ can only be achieved executing an applicable sequence of $\mathsf{apply_{\xi,\omega}}$ actions that, starting in the corresponding initial state reach a state $G \subseteq s_n$. This means that the action model used by the $\mathsf{apply_{\xi,\omega}}$ actions has to be consistent with the traversed intermediate states. We know that this is true by the definition of the $\mathsf{apply_{\xi,\omega}}$ so hence, the sub-sequence of $\mathsf{apply_{\xi,\omega}}$ appearing in $\pi'$ to solve $P'$ is a solution plan to $P=\tup{F,A[\cdot],I,G}$. 
\end{small}
\end{proof}

@@

\begin{proof}[Proof]
\begin{small}
By definition ${\mathcal I}_{\Psi,\xi}$ fully captures the set of elements that can appear in an action model $\xi$ using predicates $\Psi$. Furthermore, the compilation does not discard any possible action model definable within ${\mathcal I}_{\Psi,\xi}$. This means that for every plan $\pi$ that solves $P=\tup{F,A[\cdot],I,G}$, we can build a plan $\pi'$ selecting the appropriate actions for inserting precondition and effects to the corresponding action model and then selecting the  $\mathsf{apply_{\xi,\omega}}$ actions that allows to transform the initial state $I$ into a state $G \subseteq s_n$.
\end{small}
\end{proof}

\subsubsection{The bias of an initially {\em empty} action model}
Given that at the initial state of the classical planning compilation any $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ fluent is false, our compilation can introduce a bias to the solutions of the $P=\tup{F,A[\cdot],I,G}$ classical planning task preferring solutions that imply action models with a smaller number of preconditions and effects (i.e., a les number of {\em insert} actions).

>>>>

\begin{abstract}
This paper shows how to relax a strong assumption of the {\em plan recognition as planning} approach that is {\em knowing the action model of the observed agents}. The paper introduces a novel formulation for classical planning in a setting where no action model is given (instead, only the state variables and the action parameters are known) and it shows that this formulation neatly fits with the {\em plan recognition as planning} approach for {\em goal recognition}. The empirical evaluation evidences that this novel formulation allows to solve standard goal recognition benchmarks, still using an off-the-shelf classical planner, but without {\em a priori} knowing the action model of the observed agents.  
\end{abstract}

\section{Introduction}
\label{sec:introduction}
{\em Goal recognition} is a particular classification task in which each class represents a different goal and classification examples are observations of agents acting to achieve one of those goals. Despite there exists a wide range of different approaches for {\em goal recognition}, {\em plan recognition as planning}~\cite{ramirez2009plan,ramirez2012plan} is one of the most popular and it is currently at the core of various model-based activity recognition tasks such as, {\em goal recognition design}~\cite{KerenGK14}, {\em deceptive planning}~\cite{masters2017deceptive}, {\em planning for transparency}~\cite{macnally2018action} or {\em counter-planning}~\cite{PozancoEFB18}.

{\em Plan recognition as planning} leverages the action model of the observed agents and an off-the-shelf classical planner to compute the most likely goal of that agents. In this paper we show how to relax the assumption of {\em knowing the action model of the observed agents}, which frequently becomes a too strong assumption when applying {\em plan recognition as planning} at real-world problems.  In particular, the paper introduces a novel formulation for classical planning in a setting where no action model is given (instead, only the state variables and the action parameters are known) and it shows that this formulation neatly fits with the successful {\em plan recognition as planning} approach. The empirical evaluation evidences that this novel formulation allows to solve standard goal recognition benchmarks, still using an off-the-shelf classical planner, but without {\em a priori} knowing the action model of the observed agents.

\section{Background}
\label{sec:background}
This section formalizes the {\em planning model} we follow, the kind of {\em observations} that input the {\em goal recognition} task, and the {\em plan recognition as planning} approach for {\em goal recognition}.

@@

\begin{itemize}
\item The {\bf observed actions} are consistent with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. The number of observed actions, $l$, ranges from $0$ (fully unobserved action sequence) to $|\pi|$ (fully observed action sequence).
\item The {\bf observed states} $\tup{s_0^o, s_1^o, \ldots, s_m^o}$ is a sequence of possibly {\em partially observable states}, except for the initial state $s_0^o$, which is fully observed. A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,P)$, meaning that $\forall i, s_i^o \subseteq s_i$. The number of observed states, $m$, range from 1 (the initial state, at least), to $|\pi|+1$, and each {\em observed} states comprises $[1,|F|]$ fluents (the observation can still miss intermediate states that are {\em unobserved}).
\end{itemize}

@@

\subsection{Goal recognition with classical planning}
{\em Goal recognition} is a particular classification task in which each class represents a different possible goal $G\in G[\cdot]$ and there is a single classification example, $\mathcal{O}(\tau)$, that represents the observation of agents acting to achieve a goal $G\in G[\cdot]$.

Following the {\em naive Bayes classifier}, the {\em solution} to the {\em goal recognition} task is the subset of goals in $G[\cdot]$ that maximizes this expression.
\begin{align}
argmax_{G\in G[\cdot]} P(\mathcal{O}|G) P(G).
\end{align}

The {\em plan recognition as planning} approach shows that the $P(\mathcal{O}|G)$ likelihood can be estimated leveraging the action model of the observed agents and an off-the-shelf classical planner~\cite{ramirez2012plan}. Given a {\em classical planning problem} $P=\tup{F,A,I,G[\cdot]}$ (where $G[\cdot]$ represents the set of {\em recognizable} goals) then $P(\mathcal{O}|G)$ is estimated computing, for each goal $G\in G[\cdot]$, the cost difference of the solution plans to these two classical planning problems:
\begin{itemize}
\item $P^{\top}_G$, the classical planning problem built constraining $P=\tup{F,A,I,G}$ to achieve the particular goal $G\in G[\cdot]$ through a plan $\pi^\top$ that is {\em consistent} with the input observation $\mathcal{O}(\tau)$.
\item $P^{\bot}_G$, the classical planning problem that constrains solutions of $P=\tup{F,A,I,G}$ to plans $\pi^\bot$, that achieve $G\in G[\cdot]$, but that are {\em inconsistent} with $\mathcal{O}(\tau)$.
\end{itemize}

The higher the value of the $cost(\pi^\top)-cost(\pi^\bot)$ difference, the higher the probability of the observed agents to aim goal $G\in G[\cdot]$. With this regard, {\em plan recognition as planning} uses the {\em sigmoid function} to map the previous cost difference into a likelihood:

@@

\section{Planning with unknown domain models}
\label{sec:planning}
This section introduces a novel formulation for classical planning in a setting where no action model is given. This setting has already shown related to the learning of action models for planning~\cite{SternJ17}. In particular it can be seen as an extreme scenario when the action model is learned from a single example that contains only two state observations: the initial state and the goals. A {\em classical planning with unknown domain models} is then a tuple $P=\tup{F,A[\cdot],I,G}$, where $A[\cdot]$ is a set of actions s.t., the semantics of each action $a\in A[\cdot]$ is unknown (i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined).

A solution to this task is a sequence of actions $\pi=\tup{a_1, \ldots, a_n}$ whose execution induces a trajectory $\tau(\pi,I)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and {\em there exists} at least one possible action model (e.g. one possible definition of the $\rho$ and $\theta$ functions within the given state variables) that satisfies $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$, for every {\small $1\leq i\leq n$}, and such that the reached final state meets the goal conditions, $G \subseteq s_n$.

Next we show that the space of possible STRIPS action models can be encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this encoding to solve $P=\tup{F,A[\cdot],I,G}$ problems with an off-the-shelf classical planner and the properties of this approach.

@@

Despite any element of ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schema $\xi$, the actual space of possible \strips\ schemata is bounded by constraints of three kinds:
\begin{enumerate}
\item {\bf Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. {\em Typing constraints} are also of this kind~\cite{mcdermott1998pddl}. 
\item {\bf Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in the {\em blocksworld} one can argue that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists of an action schema $\xi$ because, in this specific domain, a block cannot be on top of itself. {\it State invariants} are also constraints of this kind~\cite{fox1998automatic}. 
\item {\bf Observation constraints}. An observation $\mathcal{O}(\tau)$ depicts {\em semantic knowledge} that constraints further the space of possible action schemata.   
\end{enumerate}

@@

In this work we introduce a propositional encoding of the {\em preconditions}, {\em negative}, and {\em positive} effects of a \strips\ action schema $\xi$ using only fluents of two kinds {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} (where $e\in{\mathcal I}_{\Psi,\xi}$). This encoding exploits the syntactic constraints of \strips\, so it is more compact that the one previously proposed by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} for learning classical planning action models. In more detail, if {\tt\small pre\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em precondition} in $\xi$. If {\tt\small pre\_e\_$\xi$} and {\tt\small eff\_e\_$\xi$} holds it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em negative effect} in $\xi$ while if $pre\_e\_\xi$ does not hold but {\tt\small eff\_e\_$\xi$} holds, it means that $e\in{\mathcal I}_{\Psi,\xi}$ is a {\em positive effect} in $\xi$. Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema with {\tt\small pre\_e\_stack} and {\tt\small eff\_e\_stack} fluents ($e\in{\mathcal I}_{\Psi,stack}$).

\subsection{A classical planning compilation for planning with unknown domain models}
Now we show how we adapt the {\em classical planning compilation for learning \strips\ action models}~\cite{aineto2018learning} to address the task of {\em planning with unknown domain models}, using our propositional encoding of STRIPS action models.

Given a classical planning problem with unknown domain models $P=\tup{F,A[\cdot],I,G}$ we create a classical planning problem $P'=\tup{F',A',I,G}$ such that:
\begin{itemize}
\item $F'$ extends $F$ with a fluent $mode_{insert}$, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. This is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding action parameters $pars(\xi)$.

\item $A'$ replaces the actions in $A$ with two types of actions.
\begin{enumerate}
\item Actions for {\em inserting} a {\em precondition}, {\em positive} effect or {\em negative} effect in $\xi$ following the syntactic constraints of \strips\ models. 
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in \Psi_{\xi}$ to the action model $\xi$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$, $eff_p$ exist in $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p}(\xi), \neg eff_{p}(\xi), mode_{insert}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p}(\xi)\}.
\end{align*}
\end{small}

@@

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p}(\xi), mode_{insert}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}

\item Actions for {\em applying} an action model $\xi$ built by the {\em insert} actions and bounded to objects $\omega\subseteq\Omega^{|pars(\xi)|}$ (where $\Omega$ is the set of {\em objects} used to induce the fluents $F$ by assigning objects in $\Omega$ to the $\Psi$ predicates and $\Omega^k$ is the $k$-th Cartesian power of $\Omega$). The action parameters, $pars(\xi)$, are bound to the objects in $\omega$ that appear in the same position.
\end{enumerate}
\end{itemize}

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\implies p(\omega)\}_{\forall p\in\Psi_\xi},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{\neg pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi}\},\\
&\{\emptyset\}\rhd\{mode_{insert}\}.
\end{align*}
\end{small}

The intuition of the compilation is that the dynamics of the actions for {\em applying} an action model $\xi$ is determined by the values of the corresponding $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ fluents in the current state. For instance, executing {\tt{\small (apply\_stack blockB blockA)}} in a state $s$ implies activating the preconditions and effects of {\tt{\small apply\_stack}} according to the values of $\{pre\_e\_stack, eff\_e\_stack\}_{\forall e\in{\mathcal I}_{\Psi,stack}}$ fluents in $s$. This means that if the current state $s$ holds $\{{\tt\scriptsize (pre\_stack\_holding\_v1),(pre\_stack\_clear\_v2)}\} \subset s$, then it must be checked that positive literals {\tt{\small (holding blockB)}} and {\tt{\small (clear blockA)}} hold in $s$. Otherwise, a different set of precondition literals will be checked for the stack action. The same applies to the positive and negative effects. Executing {\tt{\small (apply\_stack blockB blockA)}}, will add the literals {\tt{\small (on blockB blockA),(clear blockB),(not(clear blockA)),(handempty)}} and {\tt{\small(not(clear blockB))}} to the successor state only if {\tt{\small stack}} has been correctly programmed by the {\em insert} actions.

\begin{figure}[hbt!]
	{\scriptsize\tt
		{\bf 00} : (insert\_pre\_stack\_holding\_v1) \\
		01 : (insert\_pre\_stack\_clear\_v2)\\
		{\bf 02} : (insert\_eff\_stack\_clear\_v1)\\
		03 : (insert\_eff\_stack\_clear\_v2)\\
		04 : (insert\_eff\_stack\_handempty)\\
		05 : (insert\_eff\_stack\_holding\_v1)\\
		06 : (insert\_eff\_stack\_on\_v1\_v2)\\
		{\bf 07} : (apply\_stack blockA blockB)\\
	}
	\caption{\small Plan computed when solving the classical planning problem output by our compilation corresponding to a classical planning with unknown domain models.}
	\label{fig:plan-lplan}
\end{figure}

Figure~\ref{fig:plan-lplan} shows a solution plan computed when solving a $P'=\tup{F',A',I,G}$ classical planning problem output by our compilation. In the initial state of that problem the robot is holding {\tt blockA} while {\tt blockB} is clear and the single problem goal is having {\tt blockA} on top of {\tt blockB}. The plan shows that the {\em insert} actions for the action model {\tt\small stack} (steps $00-01$ insert the preconditions of the {\tt\small stack} model, steps $02-06$ insert the action model effects), and step $07$ is the plan postfix that applies the programmed action model to achieve the goals $G$. Note that another valid solution could be computed for instance, inserting the same preconditions and effects into the {\em unstack} action model and then applying the {\tt\small (unstack blockA blockB)} action instead of {\tt\small(stack blockA blockB)}.

\subsection{Compilation properties}
Now we present some properties of the compilation scheme.

@@

\begin{proof}[Proof]
\begin{small}
Once a given precondition or effect is inserted into an action model it can never be removed back and once an action model is applied it cannot be {\em reprogramed}. In the compiled problem the value of $F$, the fluents of the original problem, can exclusively be modified via $\mathsf{apply_{\xi,\omega}}$ actions.  The set of goals $G$ can only be achieved executing an applicable sequence of $\mathsf{apply_{\xi,\omega}}$ actions that, starting in the corresponding initial state reach a state $G \subseteq s_n$. This means that the action model used by the $\mathsf{apply_{\xi,\omega}}$ actions has to be consistent with the traversed intermediate states. We know that this must be true by the definition of the $\mathsf{apply_{\xi,\omega}}$ so hence, the sub-sequence of $\mathsf{apply_{\xi,\omega}}$ appearing in $\pi'$ to solve $P'$ is a solution plan to $P=\tup{F,A[\cdot],I,G}$. 
\end{small}
\end{proof}

@@

\begin{proof}[Proof]
\begin{small}
By definition ${\mathcal I}_{\Psi,\xi}$ fully captures the set of elements that can appear in an action model $\xi$ using predicates $\Psi$. Furthermore, the compilation does not discard any possible action model definable within ${\mathcal I}_{\Psi,\xi}$. This means that for every plan $\pi$ that solves $P=\tup{F,A[\cdot],I,G}$, we can build a plan $\pi'$ by selecting the appropriate actions for inserting precondition and effects to the corresponding action model and then selecting the corresponding $\mathsf{apply_{\xi,\omega}}$ actions that transform the initial state $I$ into a state $G \subseteq s_n$.
\end{small}
\end{proof}

\subsubsection{The bias of an initially {\em empty} action model}
Any $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ fluent is false at the initial state of our compilation. This fact can introduce a bias to the solutions of the $P=\tup{F,A[\cdot],I,G}$ classical planning task preferring solutions that imply action models with a smaller number of {\em preconditions}/{\em effects} (i.e., that imply a lower number of {\em insert} actions).

This bias might be eliminated defining a cost landscape where {\em insert} actions has {\em zero cost} while $\mathsf{apply_{\xi,\omega}}$ actions has a {\em positive constant cost}. In practice, since classical planners are not proficiency optimizing cost landscapes of this kind, we use a different approach that disregard the cost of actions the {\em insert} actions. Our approach is to use SAT-based planning because it can apply all the required actions for inserting preconditions in a single planning step (in parallel), because these actions do not interact. Further, actions for inserting action effects are also applied in a single planning step so the plan horizon for programming any action model is always bound to 2, which significantly reduces the planning horizon. The SAT-based planning approach is also convenient because its ability to deal with populated with dead-ends and because symmetries in the insertion of preconditions/effects into an action model do not affect to performance.
